---
title: "Statistical Rethinking 2024"
subtitle: "Week 1 lecture notes"
date: Jan 5 2024
toc: true
toc-depth: 4
number-sections: false
format:
  html:
    code-fold: true
    page-layout: full
editor: visual
reference-location: margin
---

```{r global_options, include=F, warning=F, message=F, echo=F, error=F}
library(here)
library(knitr)
source("../Methods Cordner/misc/prep.r")
```

## Difference between Bayesian and frequentist

## Bayes Theorem

## Enumerating sequences of events

### The counting rule

This is just to get our thinking started, because it's nothing other than counting. If event A could occur 3 different ways, and event B could occur two different ways, then the total number of ways events A and B can occur together is 3x2=6.

This principle generalizes to *m x n* ways for events A and B to co-occur. This principle holds for sequential events as well. For both cases, the events must be independent and not change in any way across the course of the calculations.

### Permutations

A permutation is a rearrangement of the order of a sequence. Consider the letters *a, b, c, d*. How many different ways can they be ordered?

We can brute force this.

```{r}
per <- read_excel("permutation practice.xlsx",
                  sheet="permutations")
```

### combinations

## 
