---
title: "Statistical Rethinking"
subtitle: "2026 problem sets"
date: Feb 2026
toc: true
toc-depth: 4
number-sections: false
format:
  html:
    code-fold: false
    page-layout: full
editor: visual
reference-location: margin
---

```{r global_options, include=F, warning=F, message=F, echo=F, error=F}
library(here)
library(knitr)
source("../../prep (May 2025).r")
```


## A1

To study honesty, behavioral scientists have used an experiment called the Random Allocation Game (RAG). In a RAG, participants are given a single coin. Participants flip the coin, and if the result is heads, they win a small cash prize (like 10 Euros). Participants flip the coin in private—the experimenter cannot see or verify the result, and participants know this. 

While it is impossible to know if any individual participant honestly obeys the result of the coin flip, in the aggregate the proportion of prize claims provides information about the frequency of honesty in the sample. For example, if everyone claims the prize, then probably a lot of them are liars. 

Suppose 10 participants play a RAG and 8 of them claim the prize. Using the “garden of forking data” approach, how many ways are there to realize this sample (8 out of 10), if all participants are honest? How many ways, if 5 of the participants are honest? Can you figure out the number of honest participants
that maximizes the number of ways to realize the observed sample (8 out of 10)?

### Setting up the Bayesian engine

1. Set a range of plausible values (the model space)
2. Calculate the likelihood of the data for each plausible value
3. Set the prior probability of each plausible value
4. Multiply likelihood by prior to get the numerator
5. Divide numerator by denominator to get posterior

#### Model space (hypotheses)

```{r}

honest <-seq(0, 1, by=0.1)
honest

# I believe that 70 percent are honest
prior <- c(.01, .02, .03, .06, .08, .12, .15, .18, .15, .12, .08)

prior
sum(prior)

likelihood <- dbinom(x=8,
                     size=10,
                     prob=honest)

numerator <- likelihood*prior

denominator <- sum(numerator)

posterior <- numerator / denominator

out <- data.frame(hypothesis=honest, likelihood, prior, numerator, posterior) %>%
  round(3)

out %>%
  gt() %>%
  gt_highlight_rows(rows=9) %>%
  fmt_percent(column=c(1,3),
              decimals=0) %>%
  fmt_percent(column=5,
              decimals=1) %>%
  opt_table_font(font="Gill Sans Mt",
                 size=18) %>%
  cols_align(align="center") 


```


What should we expect to see? A 50/50 chance of claiming a prize

```{r}
?rbinom

set.seed(125)

ot <- rbinom(1e3, size=10, prob=.5)

frq(ot)

```






