---
title: "Statistical Rethinking"
subtitle: "Chapter 3 notes"
author: "Dan Killian"
date: "2/13/2022"
output: 
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_depth: 3
    toc_float: true
    theme: paper
    fig.caption: true
    code_folding: show
    df_print: kable
---


```{r global_options, include=F, warning=F, message=F, echo=F, error=F}
# standard figure size and generate clean output
knitr::opts_chunk$set(fig.height=4, fig.width=6, warning=FALSE, message=FALSE, cache=TRUE, error=T, echo=T)

library(here)
source("prep.R")

```

Consider Bayes Theorem in the context of medical testing. Let there be a medical test that is either positive (+) or negative (-). 

$$Pr(infected|+)=\frac{Pr(+|infected)Pr(infected)}{Pr(+)}$$

Given that we are dealing with a binary outcome, we can rewrite the denominator as the weighted average of positive and negative tests. We'll write this in terms of probabilities. 

$$Pr(infected|+)=\frac{Pr(+|infected)Pr(infected)}{Pr(+|infected)Pr(infected) + Pr(+|not infected)Pr(not infected)}$$

Let's take the following: 

If you test positive, you're infected 95% of the time. 
If you test positive, you're not infected 1% of the time. 
0.1% of the population is infected. 

```{r}

Pr_Positive_Infected <- 0.95
Pr_Positive_NotInfected <- 0.01
Pr_Infected <- 0.001

Pr_Positive <- Pr_Positive_Infected * Pr_Infected +
  Pr_Positive_NotInfected * ( 1 - Pr_Infected )

( Pr_Infected_Positive <- Pr_Positive_Infected*Pr_Infected / Pr_Positive )

```

That is all. 

## 3.1 Sampling from a grid-approximate posterior

```{r}

p_grid <- seq( from=0 , to=1 , length.out=1000 )
prob_p <- rep( 1 , 1000 )
prob_data <- dbinom( 6 , size=9 , prob=p_grid )
posterior <- prob_data * prob_p
posterior <- posterior / sum(posterior)

ggplot(data.frame(p_grid, posterior), aes(p_grid, posterior)) + 
  geom_line()

```
Let's do the same thing with samples from the posterior. 

```{r}

## R code 3.3
samples <- data.frame(x=sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )) %>%
  rownames_to_column("id")

## R code 3.4

ggplot(samples, aes(id, x)) +
  geom_hline(yintercept=.67, color="darkblue", size=2, alpha=.7) +
  geom_point(size=.6, color="dodgerblue", alpha=.6) +
  theme(axis.ticks.x=element_blank())

```

```{r}

ggplot(samples, aes(x)) + 
geom_density(color="dodgerblue", size=1.2, fill="lightblue", alpha=.3)

```

# 3.2 Sampling to summarize

Why would we want to draw samples from the posterior distribution? We do this to conduct statistical inference. These could follow three main themes: (1) intervals of defined boundaries, and (2) questions about intervals of defined probability mass, and (3) questions about point estimates. 

### 3.2.1 Intervals of defined boundaries

What's the probability that the proportion of water on Earth is less than 50%?


Recall that the posterior is a probability distribution that sums to 1, so we can simply add up the posterior values along the grid that are less than 50%. 

```{r}
## R code 3.6
# add up posterior probability where p < 0.5
sum( posterior[ p_grid < 0.5 ] )

```
More intuitive is to work with a sample of parameter estimates, which we interpret as representing the true probability distribution. So we simply look at what proportion of estimates are less than 50%, from the entire distribution of parameter estimates. 

```{r}
## R code 3.7
sum( samples < 0.5 ) / 1e4

```
What proportion of estimates are between 50 and 75 percent? 

```{r}
## R code 3.8
sum( samples > 0.5 & samples < 0.75 ) / 1e4
```
# 3.2.2 Intervals of defined mass

Let's look at the range and endpoints defined by the probability mass, rather than a pre-defined interval of unknown mass. 

What are the boundaries of the lower 80% of the posterior distribution? 

```{r}
## R code 3.9
quantile( samples$x , 0.8 )

```

What interval defines the middle 80% of the posterior probability mass? 

```{r}

## R code 3.10
quantile( samples[,2] , c( 0.1 , 0.9 ) )

```
These intervals assign equal mass to each side, which doesn't work as well for assymetric distributions. 

```{r}
## R code 3.11
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep(1,1000)
likelihood <- dbinom( 3 , size=3 , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
samples <- data.frame(x = sample( p_grid , size=1e4 , replace=TRUE , prob=posterior ))

## R code 3.12
PI( samples , prob=0.5 )

## R code 3.13
HPDI( samples , prob=0.5 )

## R code 3.14
p_grid[ which.max(posterior) ]

## R code 3.15
chainmode( samples , adj=0.01 )

## R code 3.16
mean( samples )
median( samples )

## R code 3.17
sum( posterior*abs( 0.5 - p_grid ) )

## R code 3.18
loss <- sapply( p_grid , function(d) sum( posterior*abs( d - p_grid ) ) )

## R code 3.19
p_grid[ which.min(loss) ]

## R code 3.20
dbinom( 0:2 , size=2 , prob=0.7 )

## R code 3.21
rbinom( 1 , size=2 , prob=0.7 )

## R code 3.22
rbinom( 10 , size=2 , prob=0.7 )

## R code 3.23
dummy_w <- rbinom( 1e5 , size=2 , prob=0.7 )
table(dummy_w)/1e5

## R code 3.24
dummy_w <- rbinom( 1e5 , size=9 , prob=0.7 )
simplehist( dummy_w , xlab="dummy water count" )

## R code 3.25
w <- rbinom( 1e4 , size=9 , prob=0.6 )

## R code 3.26
w <- rbinom( 1e4 , size=9 , prob=samples )

## R code 3.27
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom( 6 , size=9 , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )

## R code 3.28
birth1 <- c(1,0,0,0,1,1,0,1,0,1,0,0,1,1,0,1,1,0,0,0,1,0,0,0,1,0,
            0,0,0,1,1,1,0,1,0,1,1,1,0,1,0,1,1,0,1,0,0,1,1,0,1,0,0,0,0,0,0,0,
            1,1,0,1,0,0,1,0,0,0,1,0,0,1,1,1,1,0,1,0,1,1,1,1,1,0,0,1,0,1,1,0,
            1,0,1,1,1,0,1,1,1,1)
birth2 <- c(0,1,0,1,0,1,1,1,0,0,1,1,1,1,1,0,0,1,1,1,0,0,1,1,1,0,
            1,1,1,0,1,1,1,0,1,0,0,1,1,1,1,0,0,1,0,1,1,1,1,1,1,1,1,1,1,1,1,1,
            1,1,1,0,1,1,0,1,1,0,1,1,1,0,0,0,0,0,0,1,0,0,0,1,1,0,0,1,0,0,1,1,
            0,0,0,1,1,1,0,0,0,0)

## R code 3.29
library(rethinking)
data(homeworkch3)

## R code 3.30
sum(birth1) + sum(birth2)

```

